---
title: 'Reproducing the visualisations using R programming language'
author:
- "Email: ravanasamuthram_krishnan.pooja@stud.hs-fresenius.de"
- 'Name: Pooja Ravanasamuthram Krishnan'
- 'ID: 400376705'
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  bookdown::html_document2:
    toc: true
    toc_depth: 2
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    fig_caption: true
  bookdown::pdf_document2:
    keep_tex: true
    toc: true
    toc_depth: '3'
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: '2'
header-includes: \usepackage{setspace}\doublespacing
fontsize: 12pt
urlcolor: blue
linkcolor: red
csl: "https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl"
editor_options:
  markdown:
    wrap: 72
---

# **Abstract**

This paper analyzes the origin and trends of economic sanctions (in
international law also referred to as restrictive measures), 1950–2022,
drawing from the Global Sanctions Database. We reproduce the graphs in
the study using the R programming language (original code provided in
Stata). The dataset needed for the graphs is however not provided, which
creates a problem for replication. The consequences of sanctions on
political goals are central to this paper. The types of graphs are line
graph, bar graph, scatter plot, stacked area chart and composite graph
showing different trend in data. As observed, despite the overall
complexity and variety of visual trends present in the data, all graphs
have similarities in the high customisation in order to further make the
data easily understandable, the use of the same colour schemes to
distinguish different categories, and the concentration of time as a
common factor as the X-axis, illustrating trends and shifts over
decades. This project demonstrates the utility of R in replicating
complex visualizations and offers insights into the process of analyzing
economic sanctions data.

# **Project Overview**

The project explores converting raw data into meaningful visualizations
through R, a statistical programming language, in terms of how well they
reflect findings in the paper. This project is also quite useful in
understanding how the methodologies were done in the study while it has
practical lessons on the challenges and fine points of data analysis and
visualization in R.This project focuses on replicating visualizations
from the paper "Economic Sanctions: Evolution, Consequences, and
Challenges." The paper discusses the history, objectives, and
effectiveness of economic sanctions, with an emphasis on their evolution
over time and their political and economic goals.

# **Significance of using R**
R is widely used in the data science job market and can be run on any operating
system, such as Windows, macOS, Linux, and many other environments. It can be
integrated with various programming languages, including C, C#, and Python. The
number of R users and those interested in learning it is significantly 
increasing, largely because it is free to use, eliminating the need for financial investment and licensing. As an open-source programming language, R is accessible
to anyone. With the growing number of users, many skilled professionals use R as
part of their profession, while others learn it out of personal interest.

# **Replication fo stata code and plot in R **
The capabilities of this language are extended by the Comprehensive R Archive Network
(CRAN), which contains numerous libraries and packages, such as ggplot2 and dplyr.
These resources can be applied in various fields, including machine learning and bioinformatics, further enhancing the versatility and utility of R.

```{r}

options(repos = c(CRAN = "https://cran.rstudio.com/"))

install.packages("dplyr")

install.packages("haven")

install.packages("ggplot2")

install.packages("reshape2")


```

To replicate the first graph, which is a stacked area graph, necessary libraries 
such as 'dplyr', 'tidyr', 'ggplot2', and 'readxl' are loaded. Since the dataset 
is in Excel format, the 'readxl' library is used as it supports older Excel file
formats like '.xls' and '.xlsx'. The Excel file is loaded, and the sheet = 1 
argument is specified to indicate that the first sheet of the Excel file should
be loaded.

```{r}


#Loading necessary libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)


#loading csv file
gsdb_data <- read_excel("C:/Users/pooja/OneDrive/Data Science- Economic Sanctions/GSDB_V3/GSDB_V3.xls", sheet = 1)

```

The column 'other_sanct' is renamed to 'other' for clarity and to prevent 
potential conflicts with the generic name 'other'. The 'objective' column is 
split into four separate columns: 'objective1', 'objective2', 'objective3', and 'objective4', with missing values filled automatically. A for loop is used to 
check which variables belong to each objective. Similarly, the 'success' column 
is also split into separate columns.


```{r}

gsdb_data <- gsdb_data %>% rename(other_sanct = other)

gsdb_data <- gsdb_data %>%
  separate(objective, into = c("objective1", "objective2", "objective3", 
  "objective4"), sep = ",", fill = "right")

objectives <- c("democracy", "destab_regime", "end_war", "human_rights", 
                "terrorism", "territorial_conflict", "prevent_war", 
                "policy_change", "other")

for (obj in objectives) {
  gsdb_data[[obj]] <- ifelse(gsdb_data$objective1 == obj | gsdb_data$objective2
                             == obj | gsdb_data$objective3 == obj | 
                               gsdb_data$objective4 == obj, 1, 0)
}

gsdb_data <- gsdb_data %>%
  separate(success, into = c("success1", "success2", "success3", "success4"),
           sep = ",", fill = "right")

success_types <- c("success_total", "success_part", "nego_settlement", "failed",
                   "ongoing")

for (succ in success_types) {
  gsdb_data[[succ]] <- ifelse(gsdb_data$success1 == succ | gsdb_data$success2 ==
                                succ | gsdb_data$success3 == succ | 
                                gsdb_data$success4 == succ, 1, 0)
}

sanct_begin <- gsdb_data %>%
  select(case_id, begin) %>%
  rename(year = begin)

sanct_end <- gsdb_data %>%
  select(case_id, end) %>%
  rename(year = end)

gsdb_case <- bind_rows(sanct_begin, sanct_end) %>%
  distinct() %>%
  arrange(case_id, year) %>%
  filter(year >= 1950) 

expanded_data <- gsdb_case %>%
  group_by(case_id) %>%
  complete(year = full_seq(year, 1)) %>%
  left_join(gsdb_data, by = "case_id")

expanded_data <- expanded_data %>%
  filter(year >= 1950)

expanded_data <- expanded_data %>%
  mutate(id_new = ifelse(year %in% sanct_begin$year[sanct_begin$case_id == 
  case_id], 1, 0))

expanded_data <- expanded_data %>%
  mutate(id = 1)

collapsed_data <- expanded_data %>%
  group_by(year) %>%
  summarise(total_cases = sum(id, na.rm = TRUE),
            new_cases = sum(id_new, na.rm = TRUE))

ggplot(collapsed_data, aes(x = year)) +
  geom_area(aes(y = total_cases, fill = "Total Cases"), alpha = 0.7) +  
  geom_area(aes(y = new_cases, fill = "New Cases"), alpha = 0.7) +      
  scale_fill_manual(values = c("Total Cases" = "blue", "New Cases" = "red")) +
  labs(y = "Number of sanction cases", x = "Years") +
  scale_y_continuous(breaks = seq(0, 400, by = 50)) +
  scale_x_continuous(breaks = seq(1950, 2022, by = 10)) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top", legend.title = element_blank())

ggsave("evol_sanct.png", width = 8, height = 6, dpi = 300)
ggsave("figure_1.pdf", width = 8, height = 6)

```

```{r}

#Loading necessary libraries
library(dplyr)
library(ggplot2)
library(readr)

gsdb_v3_case <- read_excel("C:/Users/pooja/OneDrive/Data Science- Economic Sanctions/GSDB_V3/GSDB_V3.xls", sheet = 1)

gsdb_v3_case <- gsdb_v3_case %>%
  mutate(year = begin)

collapsed_data <- gsdb_v3_case %>%
  group_by(year) %>%
  summarise(
    arms = sum(arms, na.rm = TRUE),
    military = sum(military, na.rm = TRUE),
    trade = sum(trade, na.rm = TRUE),
    financial = sum(financial, na.rm = TRUE),
    travel = sum(travel, na.rm = TRUE),
    other_sanct = sum(other, na.rm = TRUE)  # Use 'other' instead of 'other_sanct'
  )


collapsed_data <- collapsed_data %>%
  mutate(
    sum2 = arms + military,
    sum3 = arms + military + trade,
    sum4 = arms + military + trade + financial,
    sum5 = arms + military + trade + financial + travel,
    sum6 = arms + military + trade + financial + travel + other_sanct
  )

plot_data <- collapsed_data %>%
  select(year, arms, sum2, sum3, sum4, sum5, sum6) %>%
  pivot_longer(cols = c("arms", "sum2", "sum3", "sum4", "sum5", "sum6"),
               names_to = "category", values_to = "value") %>%
  mutate(category = factor(category, levels = c("arms", "sum2", "sum3", "sum4", 
                                                "sum5", "sum6")))

#defining color palette for the categories
colors <- c("blue", "maroon", "forestgreen", "darkorange", "grey", "red")

#plotting the graph
ggplot(plot_data, aes(x = year, y = value, fill = category)) +
  geom_area(alpha = 0.8, color = NA) +
  scale_fill_manual(
    values = colors,
    labels = c("Arms", "Military", "Trade", "Financial", "Travel", "Other")
  ) +
  labs(y = "Number of sanctions by type", x = "Years") +
  scale_y_continuous(breaks = seq(0, 900, by = 100), limits = c(0, 900), expand
                     = c(0, 0)) +
  scale_x_continuous(breaks = seq(1950, 2022, by = 10), expand = c(0, 0)) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 8),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 8),
    plot.background = element_rect(fill = "white", color = "white")
  ) +
  ggtitle("Number of Sanctions by Type Over Time")

#Saving the plot
ggsave("evol_type_levl.png", width = 10, height = 6, dpi = 300)
ggsave("figure_2a.pdf", width = 10, height = 6)


```
